---
title: "Simple Features Testing"
author: "Marc Weber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: yeti
    highlighted: default 
    toc: yes
    keep_md: true 
---

## Load and look at basics of simple features package
```{r, cache=TRUE}
library(devtools)
# install_github("edzer/sfr")
library(sf)
nc <- st_read(system.file("shape/nc.shp", package="sf"))
class(nc)
attr(nc, "sf_column")
methods(class = "sf")
head(nc)
```

### Download Oregon counties from Oregon Explorer data and load into simple features object
```{r, cache=TRUE, fig.keep='all', fig.show='hold', error=FALSE, message=FALSE}
library(sf)
# Get the url for zip file, download and unzip
# counties_zip <- 'http://oe.oregonexplorer.info/ExternalContent/SpatialDataforDownload/orcnty2015.zip'
# download.file(counties_zip, 'C:/users/mweber/temp/OR_counties.zip')
# unzip('C:/users/mweber/temp/OR_counties.zip')

# Now read into a simple features object in R
counties <- st_read('orcntypoly.shp')
# simple plot with base R
plot(counties[1], main='Oregon Counties', axes=TRUE)
# the data frame
head(counties[,1:5])
```


### Download Oregon cities from Oregon Explorer data and load into simple features object
```{r, cache=TRUE, fig.keep='all', fig.show='hold', error=FALSE, message=FALSE}
# cities_zip <- 'http://navigator.state.or.us/sdl/data/shapefile/m2/cities.zip'
# download.file(cities_zip, 'C:/users/mweber/temp/OR_cities.zip')
# unzip('C:/users/mweber/temp/OR_cities.zip')
cities <- st_read("cities.shp")
plot(cities[1])
```

### Construct a spatial object in R from a data frame with coordinate information - we'll use the built-in dataset 'quakes' with information on earthquakes off the coast of Fiji.  Construct spatial points sp, spatial points data frame, and then promote it to a simple features object.
```{r, cache=TRUE, fig.keep='all', fig.show='hold', error=FALSE, message=FALSE}
library(sp)
data(quakes)
head(quakes)
class(quakes)
# Data frames consist of rows of observations on columns of values for variables of interest. Create the coordinate reference system to use
llCRS <- CRS("+proj=longlat +datum=NAD83")
# now stitch together the data frame coordinate fields and the  
# projection string to createa SpatialPoints object
quakes_sp <- SpatialPoints(quakes[, c('long', 'lat')], proj4string = llCRS)
# Summary method gives a description of the spatial object in R. Summary works on pretty much all objects in R - for spatial data, gives us basic information about the projection, coordinates, and data for an sp object if it's a spatial data frame object.
summary(quakes_sp)

# we can use methods in sp library to extract certain information from objects
bbox(quakes_sp)
proj4string(quakes_sp)

# now promote the SpatialPoints to a SpatialPointsDataFrame
quakes_coords <- cbind(quakes$long, quakes$lat)
quakes_sp_df <- SpatialPointsDataFrame(quakes_coords, quakes, proj4string=llCRS, match.ID=TRUE)
summary(quakes_sp_df) # attributes folded back in
str(quakes_sp_df, max.level=2)

# Convert to simple features
quakes_sf <- st_as_sf(quakes_sp_df)
plot(quakes_sp_df[,3],cex=log(quakes_sf$depth/100), pch=21, bg=24, lwd=.4, axes=T) 
```