---
title: "Inroductory Spatial Analysis in R"
author: "Marc Weber"
output: html_document
---

Basic data structures in R can represent spatial data. All we need is some vectors with location and attribute information.
```{r}
cities <- c('Ashland','Corvallis','Bend','Portland','Newport')
longitude <- c(-122.699, -123.275, -121.313, -122.670, -124.054)
latitude <- c(42.189, 44.57, 44.061, 45.523, 44.652)
population <- c(20062,50297,61362,537557,9603)
locs <- cbind(longitude, latitude)
```

Plot locations, with size (cex) proportional to precip
```{r}
plot(locs, cex=sqrt(population*.0002), pch=20, col='red', main='Population')
text(locs, cities, pos=4)
# add a legend
breaks <- c(20000, 50000, 60000, 100000)
legend("topright", legend=breaks, pch=20, pt.cex=1+breaks/20000, col='red', bg='gray')
# Add a polygon to our map
lon <- c(-123.5, -123.5, -122.5, -122.670, -123)
lat <- c(43, 45.5, 44, 43, 43)
x <- cbind(lon, lat)
polygon(x, border='blue')
lines(x, lwd=3, col='red')
points(x, cex=2, pch=20)
```

Packages early on in R came at handling spatial data in their own way - maps package is great example - database of locational information that is quite handy.The maps package format was developed in S (R is implementation of S) - lines represented as a sequence of points separated by 'NA' values - think of as drawing with a pen, raising at NA, then lowering at a value.  Bad for associating with data since objects are only distinguished by separation with NA values.
```{r}
library(maps)
map()
map.text('county','oregon')
map.axes()
title(main="Oregon State")
```

The maptools package provides convenience function for making spatial objects from  map objects.
```{r, warning=FALSE, message=FALSE}
library(maptools)
counties <- map('county','oregon', plot=F, col='transparent',fill=TRUE)
counties$names
#strip out just the county names from items in the names vector of counties
IDs <- sapply(strsplit(counties$names, ","), function(x) x[2])
counties_sp <- map2SpatialPolygons(counties, IDs=IDs,
                                         proj4string=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"))
summary(counties_sp)
plot(counties_sp, col="grey", axes=TRUE)
```

Spatial classes provided in sp package have mostly standardized spatial data in R and provide a solid way to represent and work with spatial data in R.
```{r}
library(sp)
getClass("Spatial") #the base spatial class in sp, has a number of subclasses used for spatial objects
getClass("CRS") #a character class composed of a projection string in WKT format
getClass("SpatialPoints") #examine the structure of a couple of the subclasses in class "spatial"
getClass("SpatialLines")
getClass("SpatialPolygons")
getClass("SpatialPolygonsDataFrame")  
getClass("SpatialGrid")
```

Construt a spatial object in R from a data frame with coordinate information - we'll use the built-in dataset 'quakes' with information on earthquakes off the coast of Fiji.  After creating spatial object, promote it to a spatial data frame object.
```{r}
library(sp)
data(quakes)
head(quakes)
class(quakes)
# Data frames consist of rows of observations on columns of values for variables of interest. Create the coordinate reference system to use
llCRS <- CRS("+proj=longlat +datum=NAD83")
# now stitch together the data frame coordinate fields and the  
# projection string to createa SpatialPoints object
quakes_sp <- SpatialPoints(quakes[, c('long', 'lat')], proj4string = llCRS)
# Summary method gives a description of the spatial object in R. Summary works on pretty much all objects in R - for spatial data, gives us basic information about the projection, coordinates, and data for an sp object if it's a spatial data frame object.
summary(quakes_sp)

# We can use the generic plot function in R to produce a quick plot add axes as well:
plot(quakes_sp, axes=TRUE, col='blue', cex=.5) # axes option puts box around region 
#show state borders using maps package - very handy 
map('world','Fiji',fill=FALSE, cex = 2,add=T)

# we can use methods in sp library to extract certain information from objects
bbox(quakes_sp)
proj4string(quakes_sp)

#now promote the SpatialPoints to a SpatialPointsDataFrame
quakes_coords <- cbind(quakes$long, quakes$lat)
quakes_sp_df <- SpatialPointsDataFrame(quakes_coords, quakes, proj4string=llCRS, match.ID=TRUE)
summary(quakes_sp_df) #attributes folded back in
str(quakes_sp_df, max.level=2)
#now that we've associated data with the points, we can select and plot using the attributes of the data
plot(quakes_sp_df[quakes_sp_df$mag>5.5,],add=TRUE,col="Red") #plot just higher magnitued quakes in red
plot(quakes_sp_df[quakes_sp_df$depth > 500,],add=TRUE,col="Green") #plot just dpper quakes in green
```
Intro to raster data with raster package:
```{r, warning=FALSE, message=FALSE}
library(raster)
# create an empty raster
r <- raster(ncol=10, nrow = 10, xmx=-116,xmn=-126,ymn=42,ymx=46)
str(r, max.level=2)
r
r[] <- 1:ncell(r)
r
plot(r)

# a raster stack is a raster with multiple raster layers
r2 <- r * 50
r3 <- sqrt(r * 5)
s <- stack(r, r2, r3)
s
plot(s)
```
